Designing Types for R, Empirically

### 2020-10-12 paper/main.tex
related work

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to eventually propose changes to the
language, but we are aware that for any changes to be accepted by the user
community they must clearly benefit the language without endangering
backwards compatibility. Our goal is thus to find a compromise between
simplicity and usefulness; any proposal should cover common programming
idioms while remaining easy to learn and to use.

This paper focuses on a simpler problem than an entire type system, instead,
we limit our scope to giving types to function signatures. For this, we
design a simple type language, one that matches the R data types but omits
features such as parametric polymorphism and subtyping between user-defined
data types. We then extract type signature from execution traces of a corpus
of widely used libraries.  This allows us to see how far one can get with
the simple type language and identify limitations of the design. We validate
the robustness of the extracted type signature by implementing a contract
checker that weaves the types around their respective functions, and use a
both a large number of clients of the target packages as well as end-user code for validation. The contract
system can be used by the R community to experiment with type signatures and
replace ad hoc error checking code.

To sum up, our paper makes the following contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to instrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analysis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures and
  validated the robustness of the inferred type signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for the R programming language.
\end{itemize}

Our tools are open source and publicly available on GitHub\footnote{
\typetracer: \url{https://github.com/PRL-PRG/propagatr}, \contractr:
\url{https://github.com/PRL-PRG/contractr} }.
%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we introduce related work on extending dynamic languages
with static type systems and we give a short primer on the R programming
language.

@@

Dynamic programming languages such as Racket, JavaScript and PHP have
been extended post-facto with various static type systems.  In each
case, the type system was carefully engineered to match the salient
characteristics of the host languages and to foster a particular
programming style. For example, Typed Racket emphasizes functional
programming and supports the migration from untyped to fully typed
code~\cite{tf-popl08}, Hack~\cite{hack13} and TypeScript~\cite{BAT14}
focus on typing object-oriented features of PHP and JavaScript,
respectively. They allow users to intersperse typed and untyped code
in a fine-grained manner.  Other work looks at retrofitting types onto
Lua~\cite{typedlua2014} which account for the myriad ways Lua
programmers use tables.

@@

Recent work has gone into adding types to Python, the other eminent
data science language.  Typilus is an interesting piece of recent work
which explores using neural networks to infer types for Python
programs~\cite{allamanis2020typilus}, and Python itself added support
for type hints in Python 3.5~\cite{pythonspectypes}.  There is no
previous work on types for R, besides a short sketch of issues the
type system would need to tackle~\cite{turcotte2019towards}. We take
inspiration in the aforementioned works but focus on adapting them to
our target language.

@@

The R Project is a key tool for data analysis.  At the heart of the R
project is a \emph{vectorized, dynamic, lazy, functional, object-oriented}
programming language with a rather unusual combination of
features~\cite{ecoop12} designed to ease learning by non-programmers and
enable rapid development of new statistical methods.  The language was
designed~\citet{R96} as a successor to S~\cite{S88}.

>>>>

The goal of this paper is to gather data that can be used as input to
the process of designing a type system for R.  To focus our work, we
chose to limit the scope of investigation on the design of a language
for function type signatures.  Thus, this paper will use the collected
data to document the signature of user-defined functions.  For this,
we design a simple type language, one that matches the R data types.
We then extract type signature from execution traces of a corpus of
widely used libraries.  This allows us to see how far one can get with
the simple type language and identify limitations of the design. We
validate the robustness of the extracted type signature by
implementing a contract checker that weaves the types around their
respective functions, and use a both a large number of clients of the
target packages as well as end-user code for validation.  To sum up,
we make the following contributions:
\begin{itemize}
\item We implemented scalable and robust tooling to automatically
  extract type signatures and instrument R functions with checks
  based on their declared types.
\item We carried out a corpus analysis of \PACKAGES widely used and
  maintained packages to extract function type signatures and
  validated the robustness of the inferred type signatures against
  \NUMFILESRUN programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for R.
\item Our data and code are open source and publicly
  available.\footnote{ \url{github.com/PRL-PRG/propagatr},
  \url{github.com/PRL-PRG/contractr}, {\bf TODO add DATA link}}
\end{itemize}

%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section introduces related work and gives a short primer on R.

@@

Dynamic programming languages such as Racket, JavaScript, PHP and Lua
have been extended post factum with static type systems.  In each
case, the type system was carefully engineered to match the salient
characteristics of its host language and to foster a particular
programming style. For example, Typed Racket emphasizes functional
programming and migration from untyped to fully typed
code~\cite{tf-popl08}, Hack~\cite{hack13} and TypeScript~\cite{BAT14}
focus on object-oriented features of PHP and JavaScript,
respectively. They allow users to intersperse typed and untyped code
in a fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the
type system tried to account for the myriad ways Lua programmers use
tables.  Other languages adopted a mix of typed and untyped code by
design. In Julia, type annotations are needed for method dispatch and
performance~\cite{oopsla18a}. In Thorn, users could freely move
between typed and untyped code thanks to the addition of like
types~\cite{popl10}. Lastly C\# is an example of a statically typed
language which added a dynamic type~\cite{Bierman10}.

@@

Recent work has gone into adding types to Python, the other eminent
data science language.  Typilus is an interesting piece of recent work
which explores using neural networks to infer types for Python
programs~\cite{allamanis2020typilus}, and Python itself added support
for type hints in Python 3.5~\cite{pythonspectypes}.  There is no
previous work on types for R. We take inspiration in the
aforementioned works but focus on adapting them to our target
language.

@@

The R Project is a key tool for data analysis.  At the heart of the R
project is a vectorized, dynamic, lazy, functional, object-oriented
language with a rather unusual combination of features~\cite{ecoop12}
designed to ease learning by non-programmers and enable rapid
development of new statistical methods.  The language was
designed~\citet{R96} as a successor to S~\cite{S88}.
